Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-11-14 joint_inf.tex
Simplification and prepping for a bio future.

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

%% Useful packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{showlabels}

@@

@@

Existing consistency proofs for phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, their work motivates understanding when exactly this approximation breaks down, and what happens when we optimize branch lengths and tree topology using this approximation.

%EM clarify that we aren't doing sequence-level simulation somewhere.
In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model and imagine data being generated on the four taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the large-data limit.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known.

%EM Potential re-outline--
%
%EM 1. Describe setup with Fels and Farris trees, and how we are talking about the case of infinite data.
%EM 2. Describe fidelities
%EM 3. State theorems
%EM 4. Plots
%EM 5. Discussion, already below. This motivates a more thorough investigation under simulation.

\section{Phylogenetic maximum likelihood}

Set up mild notation needed to state likelihood

We consider the standard phylogenetic likelihood on unrooted trees \cite{Felsenstein2004}.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
We observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.

\subsection{Phylogenetic parameters}

\subsubsection{Topology}

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_blank}
\caption[short]{Farris topology $\tau_1$}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{felsenstein_blank}
\caption[short]{Felsenstein topology $\tau_2$}
\end{subfigure}
\caption{Two simple topologies}
\label{fig:farris-fels-top}
\end{figure}

Consider the fixed, binary four-tip tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}---this is commonly known as the ``Farris zone'' topology.

\subsubsection{Fidelities}

We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform (see either 8.6 of Semple and Steel \cite{Semple2003-em} or Appendix (TODO)).

\subsection{Classical maximum likelihood}

For a topology $\tau$ and branch lengths $t$, we make the usual assumption of independence between sites, obtaining the likelihood
\begin{equation}
\label{eq:full_likelihood}
%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
%das: I forgot that standard likelihoods are written with the parameters on the left.
%Since H is like a hidden variable, I went for an expectation-maximization notation, grouping them with the observations.
%It seems a little wonky, though, when we start having maximum likelihood ancestral states depending on \tau,t.
%Oh well...
L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \argmax_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\tau, t; \fullAlignment).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

\subsection{Joint maximum likelihood}

An alternative approach does away with the marginalization and directly estimates the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
One can compute the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates),
\end{equation}
then estimate the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment),
\end{equation}
using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and \eqref{eq:profile_likelihood_topology_bl} recovers the joint ML values of the unknown parameters.
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

%EM It seems straightforward to come up with an example of where branch length estimation is inconsistent. It seems like any time a state at an internal node is uncertain we will get different branch lengths by fixing it to the ML state.

\section{Inconsistency of joint inference}

\subsection{Parameter setting}

Consider the Farris topology with arbitrary branch lengths, i.e.,
$$
\tilde{t}=\{x_1,y_1,x_2,y_2,w\}
$$
with edge ordering in the order of the taxa, then the internal branch last.
We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.

\begin{lemma}
For $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ and $t=\{x,y,x,y,w\}$,
$$\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t})) = \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)).$$
\end{lemma}

@@

\subsection{Inconsistency in topology estimation}

Define $\ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t))$ as
$$
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
$$
We show the following.
\begin{theorem}
There exist $L_0(x, y)$ and $L_1(x, y)$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \le L_0(x, y),
$$
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t)) \ge L_1(x, y)
$$
and $L_1(x, y) > L_0(x, y)$.
\end{theorem}
Fig.~\ref{fig:inconsistency-farris} shows the region where $L_1(x, y) > L_0(x, y)$.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=.9\textwidth]{topology-inconsistency-inkscape}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

\subsection{Inconsistency in branch length estimation}

We will now consider the problem of branch fidelity estimation on the correct tree using joint ancestral state and branch fidelity estimation.
We will consider two settings on the Farris topology, corresponding to whether some branch fidelities are known ahead of time or not.
As above, we will assume that data is generated with two sister branches of fidelity $x$, with all other branches of length $y$.
We will show that, for a nontrivial subset of possible values for $x$ and $y$, the interior branch fidelity parameter $w'$ will be consistently overestimated as exactly equal to one instead of its true value of $y$.
That is, for the restricted case, with
$$
\hat{w}' = \arg\max_{w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x,y,w'\}; \fullAncestralSplitPartitions(\tau_1,\{x,y,w'\})),
$$
we show there are values for $x$ and $y$ where $\hat{w}'\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x',y',w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})).
$$
%EM We will show XXX... and introduce \hat{w'} here.
%das: updated

\subsubsection{Restricted case}

@@

\subsubsection{General case}

@@

\subsubsection{Empirical validation}

The region of inconsistency in Figure~\ref{fig:bl-loose-inconsistency} is potentially a conservative estimate of where $\hat{w}=1$.
To determine how conservative, we use the method of differential evolution \cite{Storn1997} to perform joint maximum likelihood estimation.
Since empirical estimates can be somewhat unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~\ref{sec:restricted-case} that cases where $x$ or $y$ are equal to zero or one can result in pathologies.
From this process we obtain Fig.~\ref{fig:bl-general-inconsistency}.

@@

\section{Discussion}

@@

@@

\newpage

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{showlabels}
\usepackage{palatino}
\usepackage{hyphenat}
\usepackage{thm-restate}

@@

% http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

@@

Existing consistency proofs for maximum-likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, their work motivates understanding when exactly this approximation breaks down, and what happens when we optimize branch lengths and tree topology using this approximation.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model and imagine data being generated on the four taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known.

\section*{Phylogenetic maximum likelihood}

We begin by fixing some notation needed to be precise about the notions of likelihood examined in this paper.
We assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
We observe $\nCols$ samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.

We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
%EM Wouldn't it be better to just parameterize these with x,y,w without primes? Also, it looks super pro if we can use tex fonts in this figure too. I love the mathjax fonts, which I think you can get https://packages.ubuntu.com/trusty/fonts/fonts-mathjax . LMK if you run into trouble. The font I use is "MathJax_Math".
\includegraphics[width=.95\textwidth]{farris_blank}
\caption[short]{Farris topology $\tau_1$}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{felsenstein_blank}
\caption[short]{Felsenstein topology $\tau_2$}
\end{subfigure}
\caption{Two simple topologies}
\label{fig:farris-fels-top}
\end{figure}

We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, which mean that the generating probabilities using the Hadamard transform have an especially simple form (see either 8.6 of Semple and Steel \cite{Semple2003-em} or Appendix (TODO)).

\subsection*{Two paths to maximum likelihood}

The standard phylogenetic likelihood approach on unrooted trees under the usual assumption of independence between sites is as follows.
For a topology $\tau$ and branch lengths $t$ the likelihood given observed ancestral states $\fullAncestralStates$ is
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
The probability $\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)$ is a product of transition probabilities determined by $\fullAlignment$, $\fullAncestralStates$, $\tau$, and $t$ \cite{Felsenstein2004}.

The typical approach to is to maximize the likelihood marginalized across ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
\end{equation}
to estimate the tree $\tau$ and branch lengths $t$.

An alternative approach used by \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully\hyp{}observed likelihood in \eqref{eq:full_likelihood}.
This is known as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
%EM Does this have to be unique? Can we just say "denote an estimate"?
We will use $\hat{\fullAncestralStates}$ to denote the estimate for $\fullAncestralStates$ obtained by maximization \eqref{eq:profile_likelihood}.
One can then estimate a topology and branch lengths using this profile likelihood:
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

\section*{Inconsistency of joint inference}

We can now state our results on inconsistency of joint inference.
All proofs are deferred to the Appendix.

%EM I think that this is a technicality. I propose that we just state that we can restrict to the parameterization shown in the figure (with certain branch lengths equal) and that a formal statement and proof is in the appendix.
Consider the Farris tree but with arbitrary branch lengths, i.e.,
$$
\tilde{t}=\{x_1,y_1,x_2,y_2,w\}
$$
with edge ordering in the order of the taxa, then the internal branch last.
We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.

\begin{lemma}
For $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ and $t=\{x,y,x,y,w\}$,
$$
\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t})) = \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)).
$$
\end{lemma}

@@

\subsection*{Inconsistency in topology estimation}

Define $\ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t))$ as
$$
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
$$
We show the following.
%EM Here's an example of restatable in action. See the end of the appendix.
\begin{restatable}{theorem}{topoInconsist}
There exist $L_0(x, y)$ and $L_1(x, y)$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \le L_0(x, y),
$$
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t; \fullAncestralSplitPartitions(\tau_2,t)) \ge L_1(x, y)
$$
and $L_1(x, y) > L_0(x, y)$.
\end{restatable}

Fig.~\ref{fig:inconsistency-farris} shows the region where $L_1(x, y) > L_0(x, y)$.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
%EM Sorry I didn't notice this before, but the font size everywhere in the plots should be at least doubled before submission. Fine if you want to wait one more cycle. The goal is that it should be the same size as the font in the text.
\includegraphics[width=.9\textwidth]{topology-inconsistency-inkscape}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

\subsection*{Inconsistency in branch length estimation}

We will now consider the problem of branch fidelity estimation on the correct tree using joint ancestral state and branch fidelity estimation.
We will consider two settings on the Farris topology, corresponding to whether some branch fidelities are known ahead of time or not.
As above, we will assume that data is generated with two sister branches of fidelity $x$, with all other branches of length $y$.
We will show that, for a nontrivial subset of possible values for $x$ and $y$, the interior branch fidelity parameter $w'$ will be consistently overestimated as exactly equal to one instead of its true value of $y$.
That is, for the restricted case, with
$$
\hat{w}' = \arg\max_{w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x,y,w'\}; \fullAncestralSplitPartitions(\tau_1,\{x,y,w'\})),
$$
we show there are values for $x$ and $y$ where $\hat{w}'\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x',y',w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})).
$$

\subsubsection*{Restricted case}

@@

\subsubsection*{General case}

@@

\subsubsection*{Empirical validation}

The region of inconsistency in Figure~\ref{fig:bl-loose-inconsistency} is potentially a conservative estimate of where $\hat{w}=1$.
To determine how conservative, we use the method of differential evolution \cite{Storn1997} to perform joint maximum likelihood estimation.
Since empirical estimates can be somewhat unstable, we perform two maximizations---one over $0 \le x',y',w' \le 1$ and one over $0 \le x',y' \le 1$ with $w'=1$---and take the value of $\hat{w}'$ with the higher likelihood.
%EM XXXX below because no bio journals have numbered sections.
We compute these maxima over a lattice in steps of $10^{-3}$ for $x$ and $y$ from $10^{-3}$ to $1-10^{-3}$---we do not include zero or one in our lattice to further stabilize the fits, recalling from Section~XXXX that cases where $x$ or $y$ are equal to zero or one can result in pathologies.
From this process we obtain Fig.~\ref{fig:bl-general-inconsistency}.

@@

\section*{Discussion}

Note that this two step thing is known in statistics as profile likelihood \cite{Murphy2000-ry}.
People have worked on its properties.

@@

\section*{Acknowledgements}
We would like to thank Richard Neher for helpful discussions.

David Shaw, Vladimir Minin, and Frederick Matsen were supported by NIH grants U19-AI117891 and R01-GM113246; David Shaw and Frederick Matsen were also supported by R01-AI12096.
The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.

@@

\newpage
\beginsupplement
